Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-04-04 paper.tex
minor notation update

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}     In this paper we will find that in a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much less than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \Omega^{-1}. 
\end{align}
To see why this should be expected, we need to compare vertical shear
with vertical buoyancy. For this preliminary discussion we consider a
vertically isothermal disk with aspect-ratio $\epsilon$,  
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$. For a protoplanetary disk, $\epsilon \sim 0.05$, $|q|\sim 1$ and $\gamma\simeq 1.4$.

For a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq \epsilon q \Omega^2\left(\frac{z}{H}\right),
\end{align}
where $H$ is the disk scale height. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

For $\gamma>1$ the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \Omega^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisified in a
thin disk, the disk would be stable.

@@

How short must $t_c$ be for vertical shear to prevail? A rough 
comparison of vertical shear to buoyancy for non-adiabatic
perturbations can be obtained by applying Eq. 47 of \cite{urpin03}, which is based on 
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk $\epsilon \ll 1$, so the destabilizing effect of 
vertical shear is rather weak because $|r\p_z\Omega|\sim\epsilon\Omega
\ll \Omega$. By comparison, the buoyancy frequency $N_z\sim
\Omega$. It follows that $t_c\lesssim O(\epsilon\Omega^{-1})$ is
required for instability.  \emph{In a thin  
  disk the thermal timescale must be shorter than the dynamical
  timescale by $O(\epsilon)\ll 1$ for the VSI to operate.}

To illustrate this point, we show in Fig. \ref{eqm_structure} the
equilibrium structure of a typical protoplanetary disk. Here $\rho_0$
is the midplane density and vertical shear rate and buoyancy
frequencies are normalized by the local Keplerian frequency. It is
clear that vertical shear is 
generally weak compared to buoyancy. Then even a small (positive)
vertical entropy gradient can significantly stabilize the VSI.

\begin{figure*}
  \includegraphics[width=\linewidth]{figures/rhoshearNz}
  \caption{Typical protoplanetary disk structure (the Minimum Mass
    Solar Nebla): density (left), vertical shear rate (middle) and
    buoyancy frequency (right).  
    \label{eqm_structure} 
  }
\end{figure*}

@@

Our central result is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c < \epsilon |q| \Omega^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. This equation
reflects Eq. \ref{bcrit_est} but there is no $z$ dependence. 
% While assumptions and simplifications will be made 
% to derive this, 
Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales the protoplanetary disk model shown in
Fig. \ref{eqm_structure}. Here $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}.

>>>>

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}     In this paper we will find that in a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much less than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \Omega_k^{-1}. 
\end{align}
where $\Omega_k$ is the Keplerian frequency. 
To see why this should be expected, we need to compare vertical shear
with vertical buoyancy. For this preliminary discussion we consider a
vertically isothermal disk with aspect-ratio $\epsilon$,  
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$. For a protoplanetary disk, $\epsilon \sim 0.05$, $|q|\sim 1$ and $\gamma\simeq 1.4$.

For a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq \epsilon q \Omega_k^2\left(\frac{z}{H}\right),
\end{align}
where $H$ is the disk scale height. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

For $\gamma>1$ the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \Omega_k^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisified in a
thin disk, the disk would be stable.

@@

How short must $t_c$ be for vertical shear to prevail? A rough 
comparison of vertical shear to buoyancy for non-adiabatic
perturbations can be obtained by applying Eq. 47 of \cite{urpin03}, which is based on 
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk $\epsilon \ll 1$, so the destabilizing effect of 
vertical shear is rather weak because $|r\p_z\Omega|\sim\epsilon\Omega_k
\ll \Omega_k$. By comparison, the buoyancy frequency $N_z\sim
\Omega_k$. It follows that $t_c\lesssim O(\epsilon\Omega_k^{-1})$ is
required for instability.  \emph{In a thin  
  disk the thermal timescale must be shorter than the dynamical
  timescale by $O(\epsilon)\ll 1$ for the VSI to operate.}

To illustrate this point, we show in Fig. \ref{eqm_structure} the
equilibrium structure of a typical protoplanetary disk. %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
It is
clear that vertical shear is 
generally weak compared to buoyancy. Then even a small (positive)
vertical entropy gradient can significantly stabilize the VSI.

\begin{figure*}
  \includegraphics[width=\linewidth]{figures/rhoshearNz}
  \caption{Typical protoplanetary disk structure (the Minimum Mass
    Solar Nebla): density (left), vertical shear rate (middle) and
    buoyancy frequency (right). Here $\rho_0$ is the midplane density.  
    \label{eqm_structure} 
  }
\end{figure*}

@@

Our central result is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c < \epsilon |q| \Omega_k^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. This equation
reflects Eq. \ref{bcrit_est} but there is no $z$ dependence. 
% While assumptions and simplifications will be made 
% to derive this, 
Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales the protoplanetary disk model shown in
Fig. \ref{eqm_structure}. Here $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}.
